Getting the Ball Rolling: Producing a Foundation Text for a Universal Corpus of the World's Languages

### 2014-03-20 docs/computEL/clean.tex
Major edits to intro
converted  minipages to figures

\section{Data source case studies}

@@

The ODIN data is easily accessible in XML format from the online database\footnote{http://odin.linguistlist.org/download}, where data for each language is saved in a separate XML file and the Interlinear Glossed Texts (IGTs) are encoded in the \texttt{<igt><example>...</example></igt>} tags. Each XML file is saved under a filename that matches the respective language code. While cleaning the data, filenames that does not adhere to ISO 639-3\footnote{http://www-01.sil.org/iso639-3} were excluded in the uniWaC compilation. [Huh? Should this really happen? - the codes are given on their website and look valid to me...]
\\ \\
\begin{minipage}{\columnwidth}

\quad 21 a.\quad o lesu mai \\
\indent \qquad\qquad 2sg return here \\
\indent \qquad\qquad `\emph{You return here.}' \\
\centerline{Figure 1: Fijian IGT from ODIN.}
\end{minipage}

@@

\begin{minipage}{\columnwidth}
\begin{lstlisting}[language=XML]
<igt>
  <example>
    <line>21 a. o lesu mai</line>
    <line>2sg return here</line>
    <line>`You return here.'</line>
  </example>
</igt>
\end{lstlisting} 
\centerline{Figure 2: Fijian IGT in ODIN's XML format.}
\end{minipage}
\\ \\
\noindent The IGTs from an academic publication usually follow the Leipzig Glossing Rules, where the first line is the source language text, the second line contains the word/morphemic equivalence and the third line is an English gloss. From the ODIN XML format, an IGT as of Figure 1 will be represented as in an XML snippet as in Figure 2. Eventually, we need to clean and extract (i) the source text `\emph{o lesu mai}' without the preceding index `\emph{a.}' and (ii) the target language gloss without the quotation marks `\emph{You return here}'.
\begin{minipage}{\columnwidth}
\begin{lstlisting}[language=XML]
<igt>
  <example>
    <line>(69) na-Na-tmi-kwalca-t 
    Yimas (Foley 1991)</line>
    <line>3sgA-1sgO-say-rise-PERF
    </line>
    <line>`She woke me up' 
    (by verbal action)</line>
  </example>
</igit>
\end{lstlisting} 
\smallskip
\centerline{Figure 3: Yimas IGT in ODIN's XML format.}
\end{minipage}
\\ \\

@@

[This hasn't been implemented, but I think we could try this: Split source and gloss lines into tokens according to whitespace. Check the source has more tokens. Count the number of hyphens and equal signs in each token (since they denote morphemes). Match the gloss line with a substring of the source line, according to numbers of hyphens and equal signs. If there's exactly one match, keep it, and discard tokens at the ends of the line. If there's more than one match, look for other punctuation in the initial and final tokens.]
\\ \\
\begin{minipage}{\columnwidth}
\begin{itemize}
\item \emph{Cleaner}: Removed (i) all heading and trailing text embedded in square or rounded brackets and (ii) heading double character token ending with bracket or fullstop.
\begin{itemize}
\item[(i)]
\begin{Verbatim}
^(?\s?\w{1,5}\s*[):.]\s*
\end{Verbatim}
\item[(ii)] 
\begin{Verbatim}
[\[\(].{1,}[\]\)]
\end{Verbatim}
\end{itemize}
\item \emph{Cleanest}: Only source lines without punctuation.
\end{itemize}
\end{minipage}
\\ \\

>>>>

\section{Data source case studies} \label{sec:case_studies}

@@

The ODIN data is easily accessible in XML format from the online database\footnote{http://odin.linguistlist.org/download}, where data for each language is saved in a separate XML file and the Interlinear Glossed Texts (IGTs) are encoded in the \texttt{<igt><example>...</example></igt>} tags. Each XML file is saved under a filename that matches the respective language code. While cleaning the data, filenames that does not adhere to ISO 639-3\footnote{http://www-01.sil.org/iso639-3} were excluded in the uniWaC compilation. [Huh? Should this really happen? - the codes are given on their website and look valid to me...]

\begin{figure}
\quad 21 a.\quad o lesu mai \\
\indent \qquad\qquad 2sg return here \\
\indent \qquad\qquad `\emph{You return here.}' \\
\caption{Fijian IGT from ODIN.} \label{fig:odin_fijian}
\end{figure}

% \begin{minipage}{\columnwidth}
% \quad 21 a.\quad o lesu mai \\
% \indent \qquad\qquad 2sg return here \\
% \indent \qquad\qquad `\emph{You return here.}' \\
% %\caption{Fijian IGT from ODIN.} 
% \end{minipage}

@@

\begin{figure}
\begin{lstlisting}[language=XML]
<igt>
  <example>
    <line>21 a. o lesu mai</line>
    <line>2sg return here</line>
    <line>`You return here.'</line>
  </example>
</igt>
\end{lstlisting} 
\caption{Fijian IGT in ODIN's XML format.} \label{fig:odin_fijian_xml}
\end{figure}

% \begin{minipage}{\columnwidth}
% \begin{lstlisting}[language=XML]
% <igt>
%   <example>
%     <line>21 a. o lesu mai</line>
%     <line>2sg return here</line>
%     <line>`You return here.'</line>
%   </example>
% </igt>
% \end{lstlisting} 
% \centerline{Figure 2: Fijian IGT in ODIN's XML format.}
% \end{minipage}

\noindent The IGTs from an academic publication usually follow the Leipzig Glossing Rules, where the first line is the source language text, the second line contains the word/morphemic equivalence and the third line is an English gloss. From the ODIN XML format, an IGT as of Figure 1 will be represented as in an XML snippet as in Figure 2. Eventually, we need to clean and extract (i) the source text `\emph{o lesu mai}' without the preceding index `\emph{a.}' and (ii) the target language gloss without the quotation marks `\emph{You return here}'.

\begin{figure}
\begin{lstlisting}[language=XML]
<igt>
  <example>
    <line>(69) na-Na-tmi-kwalca-t 
    Yimas (Foley 1991)</line>
    <line>3sgA-1sgO-say-rise-PERF
    </line>
    <line>`She woke me up' 
    (by verbal action)</line>
  </example>
</igit>
\end{lstlisting} 
\smallskip
\caption{Yimas IGT in ODIN's XML format.}\label{fig:odin_yimas_xml}
\end{figure}

% \begin{minipage}{\columnwidth}
% \begin{lstlisting}[language=XML]
% <igt>
%   <example>
%     <line>(69) na-Na-tmi-kwalca-t 
%     Yimas (Foley 1991)</line>
%     <line>3sgA-1sgO-say-rise-PERF
%     </line>
%     <line>`She woke me up' 
%     (by verbal action)</line>
%   </example>
% </igit>
% \end{lstlisting} 
% \smallskip
% \centerline{Figure 3: Yimas IGT in ODIN's XML format.}
% \end{minipage}

@@

[This hasn't been implemented, but I think we could try this: Split source and gloss lines into tokens according to whitespace. Check the source has more tokens. Count the number of hyphens and equal signs in each token (since they denote morphemes). Match the gloss line with a substring of the source line, according to numbers of hyphens and equal signs. If there's exactly one match, keep it, and discard tokens at the ends of the line. If there's more than one match, look for other punctuation in the initial and final tokens.]

\begin{figure}
\begin{itemize}
\item \emph{Cleaner}: Removed (i) all heading and trailing text embedded in square or rounded brackets and (ii) heading double character token ending with bracket or fullstop.
\begin{itemize}
\item[(i)]
\begin{Verbatim}
^(?\s?\w{1,5}\s*[):.]\s*
\end{Verbatim}
\item[(ii)] 
\begin{Verbatim}
[\[\(].{1,}[\]\)]
\end{Verbatim}
\end{itemize}
\item \emph{Cleanest}: Only source lines without punctuation.
\end{itemize}
\caption{Do we want this still as a figure?}
\end{figure}

% \begin{minipage}{\columnwidth}
% \begin{itemize}
% \item \emph{Cleaner}: Removed (i) all heading and trailing text embedded in square or rounded brackets and (ii) heading double character token ending with bracket or fullstop.
% \begin{itemize}
% \item[(i)]
% \begin{Verbatim}
% ^(?\s?\w{1,5}\s*[):.]\s*
% \end{Verbatim}
% \item[(ii)] 
% \begin{Verbatim}
% [\[\(].{1,}[\]\)]
% \end{Verbatim}
% \end{itemize}
% \item \emph{Cleanest}: Only source lines without punctuation.
% \end{itemize}
% \end{minipage}
